---
title: '**Predicting motor recovery after spinal cord injury: A Machine Learning exercise on simulated EMSCI data**'
subtitle: 'This is an ML programming exercise to apply for the Scientific Assistant position with the Biomedical Data Science Lab at ETH Zurich'
author: "Juan Carlos Ocampo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  github_document:
  word_document:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = FALSE,      
  warning = FALSE,  
  message = FALSE,   
  fig.align = 'center',
  dpi = 300,
  cache = FALSE     
)
```

# **Introduction**

This project develops a supervised learning model to predict ISNCSCI motor score at 12 months post-injury using data from the first 30 days post-injury on an EMSCI-like simulated dataset. Limitations are discussed at the end.

```{r packages, message=FALSE}
# Install and load all required packages
if (!require(pacman, quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(

  tidymodels,   # Main framework (recipes, parsnip, tune, rsample, etc.)
  stacks,       # For stacked ensemble modeling
  
  # Base Model Engines
  glmnet,       # ElasticNet
  ranger,       # Random Forest
  xgboost,      # XGBoost
  kernlab,      # Support Vector Machines (SVM)
  earth,        # MARS
  
  # Parallel Processing & Utilities
  doFuture,     # Parallel backend
  tictoc,       # To time processes
  
  # Data Manipulation
  dplyr,        # Data manipulation
  tidyr,        # Data tidying
  purrr,        # Functional programming
  tibble,       # Modern data frames}
  stringr,
  forcats,
  
  # Visualization & Tables
  ggplot2,      # Plotting
  gtsummary,    # Summary tables
  gt,           # Publication-ready tables
  broom,        # For tidy()
  viridis,
  shapviz,
  ggnewscale,
  patchwork,
  
  # Model Interpretation & Evaluation
  DALEX,        # For model explainability
  DALEXtra,     # Tidymodels integration for DALEX
  fastshap,     # SHAP values
  pdp,
  yardstick,    # For performance metrics (RMSE, R-squared, etc.)
  probably      # For calibration plots
)

print(paste("Packages loaded."))

# Set up parallel processing
total_cores <- parallel::detectCores(logical = TRUE)
num_workers <- max(1, total_cores - 2)
registerDoFuture()
plan(multisession, workers = num_workers)
print(paste("Registered", num_workers, "workers for parallel processing."))


# Set a global random seed for reproducibility
set.seed(123)

Sys.setenv(LANG = "en")
```

# **Simulated data**

The simulated data is a cohort based on descriptive statistics of EMSCI subjects reported by Bourguignon et al., 2022 (BMC Med). We aim to model realistic heterogeneity by creating several recovery phenotypes with high variance and potential plateauing that consider known SCI pathophysiology and recovery patterns.

```{r sim-data, echo=FALSE}
# Simulate EMSCI-like data

# Key principles:
# 1. 12-month outcome NOT used to generate acute trajectories
# 2. Multiple recovery phenotypes with high variance
# 3. Respects SCI pathophysiology and recovery patterns
# 4. Rapid early recovery followed by plateau

set.seed(42)

# Simulation parameters
n_patients <- 1500
n_centers <- 15

# Create realistic center distribution (power law - few large, many small centers)
center_probs <- (1 / (1:n_centers)^2)
center_probs <- center_probs / sum(center_probs)

# ------------------------------------------------------------------------------
# Step 1: Generate correlated demographic and injury characteristics
# ------------------------------------------------------------------------------

patients_static <- tibble(
  patient_id = 1:n_patients,
  
  # Center assignment (skewed distribution)
  center_id = sample(1:n_centers, n_patients, replace = TRUE, prob = center_probs),
  
  # Sex distribution (77% male, 23% female per EMSCI)
  sex = sample(c("Male", "Female"), n_patients, replace = TRUE, prob = c(0.77, 0.23)),
  
  # Generate uniform random variables for conditional sampling
  age_rand = runif(n_patients),
  cause_rand = runif(n_patients),
  nli_rand = runif(n_patients),
  ais_rand = runif(n_patients)
) %>%
  
  # Age: Bimodal distribution (young trauma, older non-trauma)
  # Using mixture of two normals to capture bimodal pattern
  mutate(
    age_at_injury = case_when(
      age_rand < 0.6 ~ pmax(9, pmin(94, rnorm(n(), 35, 12))),  # Young peak
      TRUE ~ pmax(9, pmin(94, rnorm(n(), 62, 15)))              # Older peak
    ),
    age_at_injury = floor(age_at_injury)
  ) %>%
  
  # Cause of injury (slightly different by sex)
  mutate(
    cause_of_injury = case_when(
      sex == "Male" & cause_rand <= 0.922 ~ "Traumatic",
      sex == "Male" ~ "Non-Traumatic",
      sex == "Female" & cause_rand <= 0.922 ~ "Traumatic",
      sex == "Female" ~ "Non-Traumatic"
    )
  ) %>%
  
  # Neurological level: Cervical more common, especially in older patients
  mutate(
    baseline_nli_level = case_when(
      nli_rand <= 0.53 ~ "Cervical",
      nli_rand <= 0.887 ~ "Thoracic",
      TRUE ~ "Lumbar"
    )
  ) %>%
  
  # AIS grade: Complex distribution influenced by age and level
  mutate(
    # Older patients and cervical injuries more likely to be complete
    ais_complete_prob = case_when(
      baseline_nli_level == "Cervical" & age_at_injury >= 60 ~ 0.50,
      baseline_nli_level == "Cervical" ~ 0.42,
      baseline_nli_level == "Thoracic" ~ 0.38,
      baseline_nli_level == "Lumbar" ~ 0.30
    ),
    
    baseline_ais = case_when(
      ais_rand <= ais_complete_prob ~ "A",
      ais_rand <= ais_complete_prob + 0.12 ~ "B",
      ais_rand <= ais_complete_prob + 0.12 + 0.19 ~ "C",
      TRUE ~ "D"
    )
  )

# ------------------------------------------------------------------------------
# Step 2: Generate baseline neurological scores (Day 1)
# ------------------------------------------------------------------------------

patients_static <- patients_static %>%
  mutate(
    # Motor score parameters by AIS grade (highly variable within grade)
    motor_mean = case_when(
      baseline_ais == "A" ~ 5,
      baseline_ais == "B" ~ 12,
      baseline_ais == "C" ~ 35,
      baseline_ais == "D" ~ 55
    ),
    motor_sd = case_when(
      baseline_ais == "A" ~ 4,
      baseline_ais == "B" ~ 6,
      baseline_ais == "C" ~ 12,
      baseline_ais == "D" ~ 15
    ),
    
    # Generate baseline motor score
    baseline_motor_score = pmax(0, pmin(100, floor(
      rnorm(n(), motor_mean, motor_sd)
    ))),
    
    # Adjust for paraplegia (UEMS = 50 for thoracic/lumbar)
    baseline_motor_score = case_when(
      baseline_nli_level %in% c("Thoracic", "Lumbar") ~ 
        pmin(baseline_motor_score, pmax(50, baseline_motor_score)),
      TRUE ~ baseline_motor_score
    ),
    
    # Zone of Partial Preservation (0-5 segments, critical for complete injuries)
    zpp_segments = case_when(
      baseline_ais %in% c("A", "B") ~ sample(0:5, n(), replace = TRUE, 
                                              prob = c(0.3, 0.25, 0.2, 0.15, 0.07, 0.03)),
      TRUE ~ 0  # Not applicable for incomplete injuries
    )
  ) %>%
  
  # Sensory scores: MUCH higher variability than motor (SD often > mean in real data)
  mutate(
    # Light touch baseline (correlated with motor but noisier)
    baseline_lt_score = pmax(0, pmin(112, floor(
      baseline_motor_score * 1.1 + 
        rnorm(n(), 0, 15) +  # High individual variability
        ifelse(baseline_ais == "A", -10, 0)  # Complete injuries lower
    ))),
    
    # Pin prick baseline (even more variable, often dissociates from LT)
    baseline_pp_score = pmax(0, pmin(112, floor(
      baseline_lt_score * runif(n(), 0.8, 1.2) +  # Variable relationship
        rnorm(n(), 0, 18)  # Very high variability
    )))
  )

# ------------------------------------------------------------------------------
# Step 3: Assign recovery phenotype (INDEPENDENT of outcome)
# ------------------------------------------------------------------------------

patients_static <- patients_static %>%
  mutate(
    # Recovery phenotype determines trajectory SHAPE, not final outcome
    # Three distinct groups: Marginal, Moderate, Good (especially for AIS-A)
    phenotype_rand = runif(n()),
    
    recovery_phenotype = case_when(
      # Complete injuries: Distinct subpopulations
      baseline_ais == "A" & phenotype_rand <= 0.35 ~ "marginal",
      baseline_ais == "A" & phenotype_rand <= 0.70 ~ "moderate",
      baseline_ais == "A" ~ "good",
      
      # AIS-B: Similar but shifted toward better recovery
      baseline_ais == "B" & phenotype_rand <= 0.25 ~ "marginal",
      baseline_ais == "B" & phenotype_rand <= 0.60 ~ "moderate",
      baseline_ais == "B" ~ "good",
      
      # AIS-C and D: Most show moderate-good recovery
      baseline_ais %in% c("C", "D") & phenotype_rand <= 0.15 ~ "marginal",
      baseline_ais %in% c("C", "D") & phenotype_rand <= 0.50 ~ "moderate",
      TRUE ~ "good"
    ),
    
    # Recovery rate parameter (how fast they improve in acute phase)
    # This affects trajectory steepness, NOT final outcome
    recovery_rate = case_when(
      recovery_phenotype == "marginal" ~ runif(n(), 0.05, 0.15),
      recovery_phenotype == "moderate" ~ runif(n(), 0.15, 0.35),
      recovery_phenotype == "good" ~ runif(n(), 0.35, 0.65)
    ),
    
    # Add patient-specific random effect (permanent characteristic)
    patient_random_effect = rnorm(n(), 0, 5)
  ) %>%
  
  # Clean up intermediate variables
  select(-motor_mean, -motor_sd, -age_rand, -cause_rand, -nli_rand, 
         -ais_rand, -ais_complete_prob, -phenotype_rand)

# ==============================================================================
# PART 2: LONGITUDINAL DATA SIMULATION (ACUTE PHASE: 0-30 DAYS)
# ==============================================================================

set.seed(43)

# Realistic assessment protocol
protocol_days <- c(3, 7, 14, 21, 30)

acute_longitudinal <- patients_static %>%
  select(patient_id, baseline_motor_score, baseline_lt_score, baseline_pp_score,
         baseline_ais, baseline_nli_level, recovery_phenotype, recovery_rate, 
         patient_random_effect, zpp_segments) %>%
  
  pmap_dfr(function(patient_id, baseline_motor_score, baseline_lt_score, 
                    baseline_pp_score, baseline_ais, baseline_nli_level,
                    recovery_phenotype, recovery_rate, patient_random_effect,
                    zpp_segments) {
    
    # Variable number of assessments (realistic dropout/scheduling)
    n_assessments <- sample(2:5, 1, prob = c(0.15, 0.25, 0.35, 0.25))
    
    # Always include Day 1, then sample from protocol
    days_post_injury <- sort(unique(c(
      1, 
      sample(protocol_days, n_assessments - 1, replace = FALSE)
    )))
    
    # --------------------------------------------------------------------------
    # Define acute phase potential improvement (separate from 12-month outcome)
    # --------------------------------------------------------------------------
    
    # Maximum possible acute improvement (30 days) based on baseline only
    acute_improvement_ceiling = case_when(
      baseline_ais == "A" ~ runif(1, 2, 8),      # Minimal acute recovery
      baseline_ais == "B" ~ runif(1, 5, 15),     # Modest acute recovery  
      baseline_ais == "C" ~ runif(1, 10, 25),    # Moderate acute recovery
      baseline_ais == "D" ~ runif(1, 15, 35)     # Substantial acute recovery
    )
    
    # ZPP bonus for complete injuries
    if (baseline_ais %in% c("A", "B")) {
      acute_improvement_ceiling <- acute_improvement_ceiling + zpp_segments * 1.5
    }
    
    # Level adjustment (lumbar recover more acutely)
    if (baseline_nli_level == "Lumbar") {
      acute_improvement_ceiling <- acute_improvement_ceiling * 1.3
    } else if (baseline_nli_level == "Cervical") {
      acute_improvement_ceiling <- acute_improvement_ceiling * 0.85
    }
    
    # Phenotype determines what fraction of ceiling is achieved
    phenotype_multiplier <- case_when(
      recovery_phenotype == "marginal" ~ runif(1, 0.2, 0.4),
      recovery_phenotype == "moderate" ~ runif(1, 0.5, 0.7),
      recovery_phenotype == "good" ~ runif(1, 0.75, 0.95)
    )
    
    actual_acute_improvement <- acute_improvement_ceiling * phenotype_multiplier
    
    # --------------------------------------------------------------------------
    # Generate trajectory using non-linear recovery curve
    # --------------------------------------------------------------------------
    
    # Time-dependent recovery fraction (rapid early, then plateau)
    # Using power function with exponent < 1 for diminishing returns
    time_fraction <- (days_post_injury / 30)^recovery_rate
    
    # Motor trajectory
    motor_trajectory <- baseline_motor_score + 
      actual_acute_improvement * time_fraction +
      patient_random_effect * (days_post_injury / 30) +
      rnorm(length(days_post_injury), 0, 3)  # Measurement noise
    
    # Sensory trajectories (even more variable, less consistent improvement)
    lt_improvement <- actual_acute_improvement * runif(1, 0.8, 1.3)
    lt_trajectory <- baseline_lt_score +
      lt_improvement * time_fraction +
      patient_random_effect * 0.8 * (days_post_injury / 30) +
      rnorm(length(days_post_injury), 0, 6)  # Higher sensory noise
    
    pp_improvement <- lt_improvement * runif(1, 0.7, 1.1)  # PP often dissociates
    pp_trajectory <- baseline_pp_score +
      pp_improvement * time_fraction +
      patient_random_effect * 0.7 * (days_post_injury / 30) +
      rnorm(length(days_post_injury), 0, 7)  # Highest sensory noise
    
    # CRITICAL: Enforce Day 1 = baseline (no noise at Day 1)
    motor_trajectory[1] <- baseline_motor_score
    lt_trajectory[1] <- baseline_lt_score
    pp_trajectory[1] <- baseline_pp_score
    
    # Return trajectory data
    tibble(
      patient_id = patient_id,
      days_post_injury = days_post_injury,
      motor_score = pmax(0, pmin(100, floor(motor_trajectory))),
      lt_score = pmax(0, pmin(112, floor(lt_trajectory))),
      pp_score = pmax(0, pmin(112, floor(pp_trajectory)))
    )
  })

# ==============================================================================
# PART 3: 12-MONTH OUTCOME (GENERATED INDEPENDENTLY)
# ==============================================================================

set.seed(44)  # Different seed to ensure independence

patients_static <- patients_static %>%
  mutate(
    # --------------------------------------------------------------------------
    # 12-month outcome based ONLY on baseline characteristics + NEW noise
    # NOT derived from acute trajectories or any intermediate variable
    # --------------------------------------------------------------------------
    
    # Base prediction from baseline factors
    outcome_base = 15 +  # Intercept
      
      # Baseline motor is strongest predictor
      baseline_motor_score * 0.65 +
      
      # Age effect (stronger in complete injuries)
      case_when(
        baseline_ais == "A" ~ -0.45 * (age_at_injury - 47),
        baseline_ais == "B" ~ -0.30 * (age_at_injury - 47),
        baseline_ais == "C" ~ -0.15 * (age_at_injury - 47),
        baseline_ais == "D" ~ -0.08 * (age_at_injury - 47)
      ) +
      
      # AIS grade effect (independent of motor score)
      case_when(
        baseline_ais == "A" ~ -12,
        baseline_ais == "B" ~ -5,
        baseline_ais == "C" ~ 8,
        baseline_ais == "D" ~ 18
      ) +
      
      # Level effect
      case_when(
        baseline_nli_level == "Cervical" ~ -8,
        baseline_nli_level == "Thoracic" ~ -3,
        baseline_nli_level == "Lumbar" ~ 12
      ) +
      
      # ZPP effect (only for complete injuries)
      ifelse(baseline_ais %in% c("A", "B"), zpp_segments * 2.5, 0) +
      
      # Cause effect
      ifelse(cause_of_injury == "Non-Traumatic", -7, 0) +
      
      # Sensory scores (weak independent effect)
      (baseline_lt_score - 56) * 0.05 +
      (baseline_pp_score - 56) * 0.03,
    
    # Add substantial INDEPENDENT noise (this is the key to preventing leakage)
    # Real recovery has high unexplained variance (~30-40% of variance is random)
    motor_score_12m = pmax(0, pmin(100, floor(
      outcome_base + 
        rnorm(n(), 0, 18) +  # Large random component
        patient_random_effect * 0.5  # Small carry-over from patient effect
    )))
  ) %>%
  
  # Clean up intermediate variables
  select(-outcome_base, -recovery_rate, -recovery_phenotype, 
         -patient_random_effect, -zpp_segments)

# ==============================================================================
# PART 4: COMBINE DATA
# ==============================================================================

emsci_mock_data <- patients_static %>%
  select(-baseline_motor_score, -baseline_lt_score, -baseline_pp_score) %>%
  left_join(acute_longitudinal, by = "patient_id") %>%
  select(
    patient_id, 
    center_id, 
    days_post_injury,
    motor_score,
    lt_score, 
    pp_score, 
    age_at_injury,
    sex, 
    cause_of_injury,
    baseline_ais, 
    baseline_nli_level, 
    motor_score_12m
  ) %>%
  arrange(patient_id, days_post_injury)

print("EMSCI-like cohort data simulation complete.")
print(paste("Total patients:", n_distinct(emsci_mock_data$patient_id)))
print(paste("Total observations:", nrow(emsci_mock_data)))
print(paste("Mean assessments per patient:", 
            round(nrow(emsci_mock_data) / n_distinct(emsci_mock_data$patient_id), 2)))
```

## Cohort characteristics

```{r cohort-summary-table, echo=FALSE}
# Create a cohort summary table

# Select only the baseline variables for the summary
table_1_data <- patients_static %>%
  select(
    age_at_injury,
    sex,
    cause_of_injury,
    baseline_ais,
    baseline_nli_level
  )

# Create the summary table
table_1_data %>%
  tbl_summary(
    # Stratify by AIS grade (most relevant)
    by = baseline_ais,
    label = list(
      age_at_injury ~ "Age at injury",
      sex ~ "Sex",
      cause_of_injury ~ "Cause of injury",
      baseline_nli_level ~ "Injury level"
    ),
        statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(age_at_injury ~ 1),
    missing_text = "Missing"
  ) %>%
  add_overall() %>%
  modify_spanning_header(
    c(stat_1, stat_2, stat_3, stat_4) ~ "**Baseline AIS grade**"
  ) %>%
  modify_caption("**Simulated cohort characteristics** (N=1,500)") %>%
  bold_labels() %>%
  modify_footnote(
    update = all_stat_cols() ~ "Frequencies mimic descriptive statistics reported by Bourguignon et al., 2022 (BMC Med) regarding EMSCI patient data."
  )
```

```{r longitudinal-summary-table, echo=FALSE, include=FALSE}
# Create a data frame with one row per patient
patient_assessment_counts <- acute_longitudinal %>%
  group_by(patient_id) %>%
  summarise(n_assessments = n()) %>%
  ungroup()

patient_assessment_counts %>%
  tbl_summary(
    include = c(n_assessments),
    type = list(n_assessments ~ "categorical"),
    label = list(n_assessments ~ "Assessments per patient"),
    missing = "no"
  ) %>%
  add_stat_label(label = n_assessments ~ "N (%)") %>%
  modify_caption("**Number of assessments per patient (First 30 days)**") %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

```

```{r outcome-distrib-viz, echo=FALSE, fig.width=10, fig.height=7}
# Create an outcome distribution violin plot with quartiles with jittered points for individual data

ggplot(patients_static, aes(x = baseline_ais, y = motor_score_12m, fill = baseline_ais)) +
  geom_violin(alpha = 0.6, draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(alpha = 0.25, width = 0.2, size = 1.0) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "red", shape = 18) +
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100), limits = c(0, 100)) +
  labs(
    title = "Distribution of 12-month motor scores by baseline AIS grade",
    subtitle = "Violin plots with quartiles; red diamonds indicate means",
    x = "Baseline AIS grade",
    y = "Motor score at 12 months (0-100)"
  ) +
    theme_bw() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 12), # Reduced from 14
    plot.subtitle = element_text(size = 9, color = "gray40"), # Reduced from 10
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )
```

```{r trayectories-plot, echo=FALSE, fig.width=10, fig.height=7}
# --- Robust, High-Quality Trajectory Visualization ---
# This approach avoids the geom_hex/ggnewscale conflict by using
# transparent lines (for density) and median/IQR ribbons (for tendency).

# 1. Define the AIS-specific color palette
ais_colors <- c("A" = "#E63946", "B" = "#F77F00", 
                "C" = "#FCBF49", "D" = "#06A77D")

# 2. Create a summary tibble for annotations (n per facet)
plot_summary_data <- emsci_mock_data %>%
  group_by(baseline_ais) %>%
  summarize(
    n_patients = n_distinct(patient_id),
    .groups = "drop"
  )

# 3. Create the plot
#    - We pipe the data in and create the factor in one step.
#    - All aesthetics are defined locally to be safe.
emsci_mock_data %>%
  mutate(
    baseline_ais = factor(baseline_ais, levels = c("A", "B", "C", "D"))
  ) %>%
  ggplot() +
  
  # LAYER 1: Individual Trajectories (The "Density Cloud")
  # We use a very low alpha to show the true density of paths
  geom_line(
    aes(x = days_post_injury, y = motor_score, group = patient_id),
    alpha = 0.05, # Use 0.05 - 0.1 depending on density
    color = "grey50"
  ) +
  
  # LAYER 2: Central Tendency & Variability (The "Story")
  # This is the *only* layer that uses fill and color scales.
  stat_summary(
    aes(x = days_post_injury, y = motor_score, 
        fill = baseline_ais, color = baseline_ais),
    fun.data = "median_hilow",        # Use Median
    fun.args = list(conf.int = 0.50), # 0.50 CI = Interquartile Range (IQR)
    geom = "ribbon",
    alpha = 0.3                     # A bit more alpha for the ribbon
  ) +
  stat_summary(
    aes(x = days_post_injury, y = motor_score, color = baseline_ais),
    fun = median,                     # Plot the median line
    geom = "line",
    linewidth = 1.2                   # Make the median line prominent
  ) +
  
  # LAYER 3: Annotations
  geom_label(
    data = plot_summary_data,
    aes(x = 1, y = 98, label = sprintf("n = %d patients", n_patients)),
    hjust = 0, vjust = 1, size = 3.5, fill = "white",
    fontface = "bold", label.r = unit(0, "lines")
  ) +

  # LAYER 4: Faceting & Scales
  facet_wrap(~ baseline_ais, nrow = 1) +
  
  # These scales now apply *only* to Layer 2, as intended.
  scale_color_manual(values = ais_colors, name = "Baseline AIS (Median)") +
  scale_fill_manual(values = ais_colors, name = "Baseline AIS (IQR)") +
  
  scale_y_continuous(limits = c(0, 100), breaks = c(0, 25, 50, 75, 100)) +
  scale_x_continuous(breaks = c(1, 7, 14, 21, 30), limits = c(0, 31)) +
  
  # LAYER 5: Theming & Labels
  labs(
    title = "Acute motor recovery trajectories by AIS grade (First 30 days)",
    subtitle = "Plot shows all individual trajectories (grey), median (solid line), and interquartile range (ribbon).",
    x = "Days post-injury",
    y = "Motor score (0-100)"
    ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16, margin = margin(b = 5)),
    plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 10)),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.caption = element_text(color = "gray50", hjust = 0, face = "italic"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey92"),
    strip.text = element_text(face = "bold", size = 14, color = "black"),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 10)
  )
```

# **Feature consolidation**

We pass all features to the stacked ensemble, relying on the inherent feature selection capabilities of the base learners.

```{r feature-consolidation, echo=FALSE}
# --- Data Consolidation and Robust Feature Preparation (Leakage-Free) ---

# 1. Create patient_id_global for tracking
emsci_analysis <- emsci_mock_data %>%
  mutate(patient_id_global = patient_id)

# 2. Separate outcome variable
outcomes <- emsci_analysis %>%
  select(patient_id, motor_score_12m) %>%
  distinct()

# 3. Extract time-invariant predictors
time_invariant <- emsci_analysis %>%
  select(
    patient_id,
    patient_id_global,
    center_id,
    age_at_injury,
    sex,
    cause_of_injury,
    baseline_ais,
    baseline_nli_level
  ) %>%
  distinct()

# 4. Calculate Robust Trajectory Features (Leakage-Free Engineered Features)
trajectory_features <- emsci_analysis %>%
  group_by(patient_id) %>%
  arrange(days_post_injury) %>%
  summarize(
    n_assessments = n(),
    
    baseline_motor = first(motor_score),
    latest_motor = last(motor_score),
    change_motor = latest_motor - baseline_motor,
    rate_motor = if_else(
      last(days_post_injury) > first(days_post_injury),
      change_motor / (last(days_post_injury) - first(days_post_injury)),
      0
    ),
    sd_motor = sd(motor_score, na.rm = TRUE),
    max_motor = max(motor_score),
    sd_lt = sd(lt_score, na.rm = TRUE) %>% replace_na(0),
    sd_pp = sd(pp_score, na.rm = TRUE) %>% replace_na(0),
    
    # A. Time-Weighted Score (Exponential weighting, 5-day half-life)
    # Give higher importance to more recent scores
    weighted_score = sum(motor_score * exp(- (max(days_post_injury) - days_post_injury) / 5)) / sum(exp(- (max(days_post_injury) - days_post_injury) / 5)),
    
    # B. Days Since Last Assessment
    # Reflects the 'age' of the latest information, critical for clinical models
    days_since_last_assessment = 30 - last(days_post_injury),
    
    # C. Motor Score at Day 7 (Imputed Lagged Feature)
    # Find score closest to Day 7, using linear approximation if Day 7 is missed.
    motor_d7_raw = motor_score[which.min(abs(days_post_injury - 7))],
    days_at_motor_d7_raw = days_post_injury[which.min(abs(days_post_injury - 7))],
    
    motor_d7_imputed = if_else(
      days_at_motor_d7_raw == 1,
      baseline_motor + rate_motor * (7 - 1), # Impute using baseline and average rate
      motor_d7_raw # Use the actual measurement if found near day 7
    ),
    
    # D. Trajectory Feature: Assessment Gap Flag
    # Flag if the first assessment after baseline was significantly delayed (>7 days)
    delayed_first_followup = any(days_post_injury[days_post_injury > 1] > 7),

    .groups = "drop"
  ) %>%
  select(-ends_with("_raw")) %>% 
  mutate(across(starts_with("sd_"), ~replace_na(., 0))) # Final NA check

# Combine all features into the final modeling dataset
modeling_data <- time_invariant %>%
  left_join(trajectory_features, by = "patient_id") %>%
  left_join(outcomes, by = "patient_id") %>%
  filter(!is.na(motor_score_12m)) %>%
  mutate(across(where(is.character), as.factor))

print("Modeling data created with static and trajectory features.")
```

# **Cross-validation strategy and preprocessing recipe**

We define a grouped 5-fold cross-validation using 3 repeats and a preprocessing recipe to clean the predictors.

```{r cv-strategy, echo=TRUE}
# Use 5-fold grouped CV by center, repeated 3 times for robust variance estimation
set.seed(123)
cv_splits <- group_vfold_cv(
  modeling_data,
  group = center_id,
  v = 5, # 10-fold may also work if the partitition is balanced
  repeats = 3 
)

# Create preprocessing recipe
emsci_recipe <- recipe(motor_score_12m ~ ., data = modeling_data) %>%
  update_role(patient_id, new_role = "ID") %>%
  update_role(patient_id_global, new_role = "ID") %>%
  update_role(center_id, new_role = "ID") %>%
  step_dummy(all_nominal_predictors(), one_hot = FALSE) %>% 
  step_normalize(all_numeric_predictors(), -all_of(c("patient_id", "patient_id_global", "center_id"))) %>% 
  step_integer(all_logical_predictors()) %>%
  step_zv(all_predictors())

print("Cross-validation and preprocessing recipe defined.")

```

We define a set of diverse base-learner models and computationally-efficient hyperparameter tuning grids for each.

```{r baselearner-spec, echo=TRUE}
# Define model specifications

# 1. ElasticNet (linear regularized)
set.seed(123)
elnet_spec <- linear_reg(
  penalty = tune(), 
  mixture = tune()
) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

# 2. Random Forest
set.seed(123)
rf_spec <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 2000                # Reduced for resource efficiency; increase if possible
) %>%
  set_engine("ranger", num.threads = 1) %>% 
  set_mode("regression")

# 3. XGBoost
# Early stopping and validation reduces unnecessary computation
set.seed(123)
xgb_spec <- boost_tree(
  trees = 500,                # Reduced for tuning efficiency
  tree_depth = tune(),
  learn_rate = tune(),  
  min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune()
) %>%
  set_engine(
    "xgboost", 
    nthread = 1,
    validation = 0.2,          # Early stopping uses validation split
    early_stop = 50            # Stop after 50 rounds without improvement
  ) %>%
  set_mode("regression")

# 4. SVM (Radial Basis Function kernel)
set.seed(123)
svm_spec <- svm_rbf(cost = tune(), rbf_sigma = tune()) %>%
  set_engine("kernlab") %>%
  set_mode("regression")

# 5. MARS (Multivariate Adaptive Regression Splines)
set.seed(123)
mars_spec <- mars(num_terms = tune(), prod_degree = tune()) %>%
  set_engine("earth") %>%
  set_mode("regression")

print("Base-learner specifications set.")

```

```{r baselearner-tune-spec, echo=TRUE}
# Random grids for high-dimensional or sensitive hyperparameters (ElasticNet, XGBoost, SVM)
# Regular grids for robust, fewer hyperparameters (RF, MARS)

# ElasticNet: 15-point random grid
set.seed(123)
elnet_grid <- grid_random(
  penalty(range = c(-5, 1)),
  mixture(range = c(0, 1)),
  size = 15                  # reduced for computation efficiency, increase if possible for all models
)

# Random Forest: Systematic grid (15 points)
set.seed(123)
rf_grid <- grid_regular(
  mtry(range = c(5, 20)),
  min_n(range = c(5, 15)),
  levels = c(mtry = 5, min_n = 3)
)

# XGBoost: 15-point random grid
set.seed(123)
xgb_grid <- grid_random(
  tree_depth(range = c(3, 10)),
  min_n(range = c(2, 20)),
  learn_rate(range = c(-3, -0.5)), # Tuning log10(learning_rate) from 0.001 to 0.3
  loss_reduction(range = c(-5, 1)),
  sample_prop(range = c(0.5, 1.0)),
  size = 15
)

# SVM: 15-point random grid
set.seed(123)
svm_grid <- grid_random(
  cost(range = c(-5, 5)),
  rbf_sigma(range = c(-5, 0)),
  size = 15
)

# MARS: Systematic grid (15 points)
set.seed(123)
mars_grid <- grid_regular(
  num_terms(range = c(5, 25)),
  prod_degree(range = c(1L, 2L)),
  levels = c(num_terms = 5, prod_degree = 3)
)

print("Hyperparameter grids finalized.")
```

We use a workflow set to tune all five models across our grouped CV splits and save the out-of-fold predictions to be used inputs for the final ensemble.

```{r baselearners-tune-run, echo=FALSE}
# Tune all base learners and blend the stack using 5-fold grouped CV
tic("Base learners tuning")

# Use workflow_set for a cleaner, more efficient workflow
set.seed(123)
emsci_workflow_set <- workflow_set(
  preproc = list(recipe = emsci_recipe),
  models = list(
    elnet = elnet_spec,
    rf = rf_spec,
    xgb = xgb_spec,
    svm = svm_spec,
    mars = mars_spec
  )
)

# Add the specific grids to the workflow_set
emsci_workflow_set <- emsci_workflow_set %>%
  option_add(grid = elnet_grid, id = "recipe_elnet") %>%
  option_add(grid = rf_grid, id = "recipe_rf") %>%
  option_add(grid = xgb_grid, id = "recipe_xgb") %>%
  option_add(grid = svm_grid, id = "recipe_svm") %>%
  option_add(grid = mars_grid, id = "recipe_mars")

# Define control settings for tuning (to save predictions for stacking)
ctrl_stack <- control_stack_grid()

# Run tuning in parallel
set.seed(123)
model_set_results <-
  workflow_map(
    emsci_workflow_set,
    fn = "tune_grid", 
    resamples = cv_splits,
    metrics = metric_set(rmse, mae, rsq),
    control = ctrl_stack,
    verbose = TRUE
  )
toc()

print("Base learners tuned.")
```

# **Stacked ensemble creation**

We create the stacked ensemble by adding the tuned base learners as candidates and blending their out-of-fold predictions using a non-negative LASSO meta-learner to find optimal, interpretable weights. Then, we fit the complete stacked model to the full training set.

```{r metalearner-fit, echo=TRUE}
# Create and fit the stacked ensemble
tic("Stacking")

# Initialize stack
set.seed(123)
emsci_stack <- stacks() %>%
  add_candidates(model_set_results) # Add all tuned models at once

print(paste("Stack initialized with", ncol(emsci_stack), "candidate models"))

# Blend the stack (fit meta-learner with non-negative LASSO)
set.seed(123)
emsci_stack_blended <- emsci_stack %>%
  blend_predictions(
    penalty = 10^seq(-4, -0.5, length.out = 20),
    mixture = 1,  # LASSO
    non_negative = TRUE,                # enforce interpretability
    metric = metric_set(rmse)           # RMSE prioritizes predictive accuracy
  )

print("Stack blended successfully")

# Fit the final stacked model on the full training data
set.seed(123)
emsci_stack_fitted <- emsci_stack_blended %>%
  fit_members()

toc()

print("Final stacked ensemble fitted.")
```

# **Meta-learner evaluation (two-stage CV)**

We perform a two-stage evaluation using an independent set of 5-fold grouped CV splits. We use the base-learner predictions as inputs and the optimal non-negative LASSO penalty found during blending and generate the final out-of-sample performance metrics.

```{r eval-prep, echo=TRUE}
# Obtain out-of-fold predictions and performance metrics
tic("Model evaluation")

# Create the L1 dataset (base learner predictions + IDs)
l1_data <- emsci_stack %>%
  mutate(.row = row_number()) %>%
  left_join(
    modeling_data %>% 
      mutate(.row = row_number()) %>% 
      select(.row, patient_id_global, center_id),
    by = ".row"
  )

print(paste("L1 dataset created:", nrow(l1_data), "patients"))

# Define NEW, INDEPENDENT CV folds for meta-learner evaluation
set.seed(456) # Use a different seed for independence
l1_folds <- group_vfold_cv(l1_data, group = center_id, v = 5)

# Get optimal penalty from Stage 1
best_penalty <- emsci_stack_blended$penalty$penalty

print(paste("Using meta-learner penalty from Stage 1:", best_penalty))

# Define meta-learner spec
# Use !!  to inject the value, prevents silent fit_resamples scope error
meta_spec <- linear_reg(
  penalty = !!best_penalty,
  mixture = 1 # LASSO
) %>%
  set_engine("glmnet", lower.limits = 0) # Non-negative 

# Define meta-learner recipe
meta_recipe <- recipe(motor_score_12m ~ ., data = l1_data) %>%
  update_role(center_id, new_role = "ID") %>%
  update_role(patient_id_global, new_role = "ID") %>%
  update_role(.row, new_role = "ID")

# Create workflow
eval_metrics <- metric_set(rmse, rsq, mae)
meta_wf <- workflow() %>%
  add_recipe(meta_recipe) %>%
  add_model(meta_spec)

# Evaluate with CV
set.seed(456)
tic("Meta-learner CV")
meta_cv_results <- fit_resamples(
  meta_wf,
  resamples = l1_folds,
  metrics = eval_metrics, # Pass the metrics
  control = control_resamples(save_pred = TRUE)
)
toc()
print("Meta-learner CV complete")

# Extract CV predictions
cv_predictions <- tidyr::unnest(meta_cv_results, .predictions)

# Create final plotting data
plot_data_cv <- cv_predictions %>%
  left_join(
    modeling_data %>% 
      mutate(.row = row_number()) %>%
      select(.row, patient_id_global, baseline_ais),
    by = ".row"
  ) %>%
  select(patient_id_global, motor_score_12m, .pred, baseline_ais)

# Calculate final CV metrics
final_cv_metrics <- collect_metrics(meta_cv_results)

print(paste("Out-of-fold predictions extracted for", 
            nrow(plot_data_cv), "patients"))
print("Performance metrics and predictions ready.")
```

## Meta-learner weights

We extract the non-negative LASSO coefficients from the fitted stack, group them by base-learner type, and visualize the total weight and number of members each model contributes to the final ensemble.

```{r metalearner-weights, fig.width=10, fig.height=7}
# Calculate percentages and prepare for plotting

# Extract meta-learner coefficients and convert model to tibble
meta_weights <- emsci_stack_fitted$coefs %>%
  broom::tidy() %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Parse model type from term name
    model_type = case_when(
      stringr::str_detect(term, "elnet") ~ "ElasticNet",
      stringr::str_detect(term, "rf") ~ "Random Forest",
      stringr::str_detect(term, "xgb") ~ "XGBoost",
      stringr::str_detect(term, "svm") ~ "SVM",
      stringr::str_detect(term, "mars") ~ "MARS",
      TRUE ~ "Other"
    ),
    # Only keep members with a positive weight
    estimate = ifelse(estimate > 0, estimate, 0)
  ) %>%
  filter(estimate > 0) %>%
  arrange(desc(estimate))

# Calculate total weight by model type (This creates the missing object)
weights_by_model <- meta_weights %>%
  group_by(model_type) %>%
  summarize(
    total_weight = sum(estimate),
    n_members = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(total_weight))

weights_plot_data <- weights_by_model %>%
  mutate(
    pct = 100 * total_weight / sum(total_weight),
    label_text = sprintf("%.1f%%\n%d members", pct, n_members)
  )

# Create enhanced weights plot
weights_plot <- ggplot(weights_plot_data, 
                       aes(x = reorder(model_type, total_weight), 
                           y = total_weight, 
                           fill = model_type)) +
  geom_col(alpha = 0.85, width = 0.7) +
  # Add value labels
  geom_text(aes(label = label_text),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  coord_flip(ylim = c(0, max(weights_plot_data$total_weight) * 1.2)) +
  # Distinct colors for each model type
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(
    title = "Meta-learner weight distribution by base model type",
    subtitle = sprintf("Non-negative LASSO selected %d of %d candidate models. Percentages show proportion of total ensemble weight.
Member count shows how many configurations were retained. Higher weight = greater contribution to final predictions.", 
                      sum(weights_plot_data$n_members),
                      ncol(emsci_stack) - 1),
    x = "Base model type",
    y = "Total weight (Sum of LASSO coefficients)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40", margin = margin(b = 10)),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 0),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text.y = element_text(size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

weights_plot
```

## Evaluation metrics

We gather performance metrics for the final stacked ensemble and the best-performing base learners.

```{r eval-metrics-table}
# --- 1. Get Unbiased Stack Metrics ---
# (This section is correct and unchanged)
stack_metrics_formatted <- final_cv_metrics %>%
  filter(.metric %in% c("rmse", "rsq", "mae")) %>%
  mutate(
    ci_lower = mean - 1.96 * std_err,
    ci_upper = mean + 1.96 * std_err,
    `Mean (95% CI)` = case_when(
      .metric == "rsq" ~ sprintf("%.3f [%.3f, %.3f]", mean, ci_lower, ci_upper),
      TRUE ~ sprintf("%.2f [%.2f, %.2f]", mean, ci_lower, ci_upper)
    ),
    Metric = case_when(
      .metric == "mae" ~ "MAE",
      .metric == "rmse" ~ "RMSE",
      .metric == "rsq" ~ "R-squared"
    )
  ) %>%
  select(Metric, `Mean (95% CI)`) %>%
  mutate(Model = "Stacked ensemble") # Add a model name for merging

# --- 2. Get Biased Base Learner Metrics ---
# (This section is correct and unchanged)
base_learners_formatted <- model_set_results %>%
  rank_results(select_best = TRUE) %>%
  filter(.metric %in% c("rmse", "rsq", "mae")) %>%
  select(wflow_id, .metric, mean, std_err) %>%
  mutate(
    Model = case_when(
      stringr::str_detect(wflow_id, "elnet") ~ "Best ElasticNet",
      stringr::str_detect(wflow_id, "rf") ~ "Best Random Forest",
      stringr::str_detect(wflow_id, "xgb") ~ "Best XGBoost",
      stringr::str_detect(wflow_id, "svm") ~ "Best SVM",
      stringr::str_detect(wflow_id, "mars") ~ "Best MARS"
    ),
    ci_lower = mean - 1.96 * std_err,
    ci_upper = mean + 1.96 * std_err,
    `Mean (95% CI)` = case_when(
      .metric == "rsq" ~ sprintf("%.3f [%.3f, %.3f]", mean, ci_lower, ci_upper),
      TRUE ~ sprintf("%.2f [%.2f, %.2f]", mean, ci_lower, ci_upper)
    ),
    Metric = case_when(
      .metric == "mae" ~ "MAE",
      .metric == "rmse" ~ "RMSE",
      .metric == "rsq" ~ "R-squared"
    )
  ) %>%
  select(Model, Metric, `Mean (95% CI)`)

# --- 3. Combine and Format for Table ---
# Get the number of models retained in the final stack
n_retained_models <- emsci_stack_blended$coefs %>%
  broom::tidy() %>%
  filter(term != "(Intercept)", estimate > 0) %>%
  nrow()

# Bind the two tibbles together
full_performance_table <- bind_rows(base_learners_formatted, stack_metrics_formatted) %>%
  # Pivot to put metrics as columns
  pivot_wider(names_from = Metric, values_from = `Mean (95% CI)`) %>%
  
  # --- FIX: The levels here MUST exactly match the strings created above ---
  mutate(Model = factor(Model, levels = c(
    "Best ElasticNet",
    "Best Random Forest",
    "Best XGBoost",
    "Best SVM",
    "Best MARS",
    "Stacked ensemble" # <-- Must match "Stacked Ensemble (Unbiased)"
  ))) %>%
  
  # This arrange call will now work correctly, using the factor levels
  arrange(Model)

# --- 4. Create gt Table ---
# (This section is correct and unchanged)
performance_gt <- full_performance_table %>%
  gt() %>%
  tab_header(
    title = "Cross-validated performance metrics",
    subtitle = paste("Stacked ensemble (retained", n_retained_models, "candidates) vs. best base learners")
  ) %>%
  cols_label(
    Model = "Model"
  ) %>%
  tab_footnote(
    footnote = "Base learner metrics are optimistically biased (from stage 1 tuning). Stacked ensemble metrics are unbiased (from independent stage 2 CV).",
    locations = cells_column_labels(columns = Model)
  ) %>%
  # This conditional style will also work now
  tab_style(
    style = list(
      cell_fill(color = "#E8F8F5"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Model == "Stacked Ensemble"
    )
  )

performance_gt
```

## Calibration plot

We assess model calibration using the out-of-fold predictions via predicted vs observed scores.

```{r calibration-plot, fig.width=10, fig.height=7}
# Calculate calibration metrics
calibration_metrics <- plot_data_cv %>%
  summarize(
    calibration_slope = coef(lm(motor_score_12m ~ .pred))[2],
    calibration_intercept = coef(lm(motor_score_12m ~ .pred))[1],
    pearson_r = cor(motor_score_12m, .pred),
    rsq = pearson_r^2
  )

# Create enhanced calibration plot
calibration_plot <- ggplot(plot_data_cv, aes(x = .pred, y = motor_score_12m)) +
  # Add reference line (perfect calibration)
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "gray40", linewidth = 0.8) +
  # Add LOESS smoother to detect systematic bias
  geom_smooth(method = "loess", se = TRUE, color = "#2E86AB", 
              fill = "#A9D6E5", alpha = 0.3, linewidth = 1) +
  # Add linear regression line
  geom_smooth(method = "lm", se = FALSE, color = "#E63946", 
              linetype = "solid", linewidth = 0.8) +
  # Points colored by AIS grade for clinical context
  geom_point(aes(color = baseline_ais), alpha = 0.5, size = 2) +
  # Add comprehensive annotation box
  annotate("text", 
           x = min(plot_data_cv$.pred) + 8, 
           y = max(plot_data_cv$motor_score_12m) - 8,
           label = sprintf(
             "Calibration slope: %.3f\nIntercept: %.3f\nR-squared: %.3f\nPearson r: %.3f", 
             calibration_metrics$calibration_slope,
             calibration_metrics$calibration_intercept,
             calibration_metrics$rsq,
             calibration_metrics$pearson_r
           ),
           hjust = 0, vjust = 1, size = 3.5,
           fontface = "bold",
           color = "gray20") +
  # Color scale for AIS grades
  scale_color_brewer(palette = "Set1", name = "Baseline AIS") +
  # Labels with clear interpretation guidance
  labs(
    title = "Calibration plot: Predicted vs. observed motor scores",
    subtitle = "Points colored by baseline AIS grade. Red line = linear fit; Blue = LOESS smoother; Dashed = perfect calibration",
    x = "Predicted motor score at 12 months",
    y = "Observed motor score at 12 months"
  ) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40", margin = margin(b = 10)),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 0),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

print(calibration_plot)
```

## Calibration by score range

We stratify the data by predicted score range to analyze calibration-in-the-large.

```{r calibration-by-range-plot, fig.width=10, fig.height=7}
# Analyze calibration stratified by predicted score range
# This reveals if model is well-calibrated across the outcome spectrum

calibration_by_range <- plot_data_cv %>%
  mutate(
    pred_category = cut(.pred, 
                       breaks = c(-Inf, 25, 50, 75, Inf),
                       labels = c("0-25", "26-50", "51-75", "76-100"))
  ) %>%
  group_by(pred_category) %>%
  summarize(
    n = n(),
    mean_predicted = mean(.pred),
    mean_observed = mean(motor_score_12m),
    calibration_error = mean_observed - mean_predicted,
    .groups = "drop"
  )

# Create calibration-in-the-large plot
cal_range_plot <- ggplot(calibration_by_range, 
                         aes(x = mean_predicted, y = mean_observed)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(aes(size = n), color = "#2E86AB", alpha = 0.7) +
  geom_segment(aes(xend = mean_predicted, yend = mean_predicted),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "#E63946", linewidth = 1) +
  geom_text(aes(label = sprintf("n=%d\nΔ=%.1f", n, calibration_error)),
            vjust = -1.5, size = 3) +
  scale_size_continuous(range = c(5, 15), name = "Sample Size") +
  labs(
    title = "Calibration-in-the-Large by predicted score range",
    subtitle = "Red arrows show calibration error (observed - predicted). Well-calibrated models have short arrows",
    x = "Mean predicted score",
    y = "Mean observed score"
  ) +
  coord_fixed() +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40", margin = margin(b = 10))
  )

cal_range_plot
```

```{r calibration-by-range-table}
# Print calibration table
cal_range_gt <- calibration_by_range %>%
  mutate(
    `Score Range` = as.character(pred_category),
    `N` = n,
    `Mean Predicted` = sprintf("%.1f", mean_predicted),
    `Mean Observed` = sprintf("%.1f", mean_observed),
    `Calibration Error` = sprintf("%.2f", calibration_error)
  ) %>%
  select(`Score Range`, N, `Mean Predicted`, `Mean Observed`, `Calibration Error`) %>%
  gt() %>%
  tab_header(
    title = "Calibration stratified by score range"
  ) %>%
  tab_footnote(
    footnote = "Calibration error = Observed - predicted. positive = underprediction",
    locations = cells_column_labels(columns = `Calibration Error`)
  )

cal_range_gt
```

## Residual diagnostics

We check the model error behavior by generating standard diagnostic plots from the out-of-fold predictions.

```{r res-diag, fig.width=10, fig.height=7}
# --- Methodologically Flawless Residual Diagnostics ---
# This chunk generates the three core diagnostic plots to 
# understand the model's error behavior.

# Ensure the 'residual' column exists on our OOF data object
# (This is already created in your 'blandaltman-plot' chunk, 
#  but we include it here for robustness)
if (!"residual" %in% names(plot_data_cv)) {
  plot_data_cv <- plot_data_cv %>%
    mutate(residual = motor_score_12m - .pred)
}

# 1. Residuals vs. Fitted Plot
# PURPOSE: To detect systematic bias and non-linear patterns.
# INTERPRETATION: A flat line at 0 is perfect. 
#   A curve = systematic bias (e.g., under-predicting high values).
p_resid_fitted <- ggplot(plot_data_cv, aes(x = .pred, y = residual)) +
  geom_point(aes(color = baseline_ais), alpha = 0.4, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "loess", color = "#E63946", se = TRUE, 
              fill = "#E63946", alpha = 0.1) +
  scale_color_brewer(palette = "Set1", name = "Baseline AIS") +
  labs(
    title = "Residuals vs. fitted values",
    subtitle = "Diagnostic for systematic bias and non-linearity",
    x = "Fitted (predicted) Values",
    y = "Residual (observed - predicted)"
  ) +
 theme_minimal() +
  theme(
    theme(legend.position = "bottom"),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    axis.title = element_text(size = 11)
  )


# 2. Normal Q-Q Plot
# PURPOSE: To describe the *character* of the errors.
# INTERPRETATION: Are errors normal (on the line) or do 
#   we have "fat tails" (S-shape) = more extreme errors?
p_qq <- ggplot(plot_data_cv, aes(sample = residual)) +
  stat_qq(aes(color = baseline_ais), alpha = 0.4) +
  stat_qq_line(color = "#E63946", linewidth = 1) +
  scale_color_brewer(palette = "Set1", name = "Baseline AIS") +
  labs(
    title = "Normal Q-Q plot",
    subtitle = "Diagnostic for character of the error distribution",
    x = "Theoretical quantiles (normal)",
    y = "Sample quantiles (residuals)"
  ) +
 theme_minimal() +
  theme(
    theme(legend.position = "bottom"),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    axis.title = element_text(size = 11)
  )

# 3. Scale-Location Plot
# PURPOSE: To detect heteroscedasticity (non-constant error).
# INTERPRETATION: A flat line is ideal (homoscedasticity). 
#   A sloped line = error magnitude depends on the predicted value.
p_scale_loc <- ggplot(plot_data_cv, aes(x = .pred, y = sqrt(abs(residual)))) +
  geom_point(aes(color = baseline_ais), alpha = 0.4, size = 1.5) +
  geom_smooth(method = "loess", color = "#E63946", se = TRUE, 
              fill = "#E63946", alpha = 0.1) +
  scale_color_brewer(palette = "Set1", name = "Baseline AIS") +
  labs(
    title = "Scale-Location plot",
    subtitle = "Diagnostic for: Heteroscedasticity (non-constant error variance)",
    x = "Fitted (predicted) Values",
    y = "√|Standardized residuals|"
  ) +
 theme_minimal() +
  theme(
    theme(legend.position = "bottom"),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    axis.title = element_text(size = 11)
  )

# Combine all plots
(p_resid_fitted) / (p_qq | p_scale_loc) + 
  plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

## Bland-Altman plot

We assess the agreement between observed and predicted scores, visualizing the mean bias and 95% limits of agreement to check for any systematic error trends across the range of scores.

```{r blandaltman-plot, fig.width=10, fig.height=7}

plot_data_cv <- plot_data_cv %>%
  mutate(
    residual = motor_score_12m - .pred,
    abs_residual = abs(residual)
  )

# Calculate Bland-Altman statistics
ba_stats <- plot_data_cv %>%
  summarize(
    mean_diff = mean(residual),
    sd_diff = sd(residual),
    loa_upper = mean_diff + 1.96 * sd_diff,
    loa_lower = mean_diff - 1.96 * sd_diff
  )

# Create Bland-Altman plot
ba_plot <- plot_data_cv %>%
  mutate(mean_score = (motor_score_12m + .pred) / 2) %>%
  ggplot(aes(x = mean_score, y = residual)) +
  # Add reference line at zero (no bias)
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.8) +
  # Add mean bias line
  geom_hline(yintercept = ba_stats$mean_diff, color = "#E63946", 
             linewidth = 0.8) +
  # Add limits of agreement
  geom_hline(yintercept = ba_stats$loa_upper, color = "#2E86AB", 
             linetype = "dotted", linewidth = 0.8) +
  geom_hline(yintercept = ba_stats$loa_lower, color = "#2E86AB", 
             linetype = "dotted", linewidth = 0.8) +
  # Add LOESS smoother (global)
  geom_smooth(method = "loess", se = TRUE, color = "#06A77D", 
              fill = "#D0F4DE", alpha = 0.3, linewidth = 1) +
  
  # FIX: Map color to baseline_ais inside aes()
  geom_point(aes(color = baseline_ais), alpha = 0.4, size = 1.5) +
  
  # Add annotations
  annotate("text", x = max(plot_data_cv$mean_score) * 0.95, 
           y = ba_stats$mean_diff, 
           label = sprintf("Mean bias: %.2f", ba_stats$mean_diff),
           hjust = 1, vjust = -0.5, size = 3.5, color = "#E63946") +
  annotate("text", x = max(plot_data_cv$mean_score) * 0.95, 
           y = ba_stats$loa_upper,
           label = sprintf("+1.96 SD: %.2f", ba_stats$loa_upper),
           hjust = 1, vjust = -0.5, size = 3.5, color = "#2E86AB") +
  annotate("text", x = max(plot_data_cv$mean_score) * 0.95, 
           y = ba_stats$loa_lower,
           label = sprintf("-1.96 SD: %.2f", ba_stats$loa_lower),
           hjust = 1, vjust = 1.5, size = 3.5, color = "#2E86AB") +
           
  # FIX: Add a color scale
  scale_color_brewer(palette = "Set1", name = "Baseline AIS") +
  
  # Labels and theme
  labs(
    title = "Bland-Altman plot: agreement between predicted and observed scores",
    subtitle = "Points colored by baseline AIS grade. Green LOESS shows global bias trend.
Red line is mean bias; Blue dotted lines are 95% limits of agreement.",
    x = "Mean of predicted and observed score",
    y = "Difference (observed - predicted)"
    ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    axis.title = element_text(size = 11)
  )

print(ba_plot)
```

## Subgroup analysis by AIS grade

We visualize the model performance stratified by the baseline AIS grade to compare each group's error against the overall model performance.


```{r eval-subgroup}
# Calculate performance metrics by AIS grade
subgroup_metrics <- plot_data_cv %>%
  group_by(baseline_ais) %>%
  summarize(
    n = n(),
    rmse = rmse_vec(truth = motor_score_12m, estimate = .pred),
    mae = mae_vec(truth = motor_score_12m, estimate = .pred),
    rsq = rsq_vec(truth = motor_score_12m, estimate = .pred),
    .groups = "drop"
  ) %>%
  arrange(baseline_ais)

# Create subgroup table
subgroup_gt <- subgroup_metrics %>%
  gt() %>%
  tab_header(
    title = "Subgroup analysis: Performance by baseline AIS grade"
  )
```

```{r eval-subgroup-plot, fig.width=11, fig.height=7}
# Calculate overall performance for reference
overall_rmse <- final_cv_metrics %>% 
  filter(.metric == "rmse") %>% 
  pull(mean)

# Prepare subgroup data for visualization
subgroup_plot_data <- subgroup_metrics %>%
  mutate(
    ais_label = sprintf("AIS %s\n(n=%d)", baseline_ais, n)
  )

# Create enhanced subgroup plot
subgroup_plot <- subgroup_plot_data %>%
  ggplot(aes(x = reorder(ais_label, -rmse), y = rmse, fill = baseline_ais)) +
  geom_col(alpha = 0.85, width = 0.7) +
  # Add reference line for overall performance
  geom_hline(yintercept = overall_rmse, linetype = "dashed", 
             color = "gray30", linewidth = 0.8) +
  # Add value labels on bars
  geom_text(aes(label = sprintf("RMSE: %.2f\nR-squared: %.3f", rmse, rsq)),
            vjust = -0.3, size = 3.5, fontface = "bold") +
  # Add annotation for reference line
  annotate("text", x = 0.7, y = overall_rmse * 1.05,
           label = sprintf("Overall RMSE: %.2f", overall_rmse),
           hjust = 0, size = 3.5, color = "gray30") +
  # AIS-specific color palette
  scale_fill_manual(
    values = c("A" = "#E63946", "B" = "#F77F00", 
               "C" = "#FCBF49", "D" = "#06A77D"),
    guide = "none"
  ) +
  labs(
    title = "Subgroup analysis: Model performance by baseline AIS grade",
    subtitle = "RMSE and R-squared stratified by injury severity. Lower RMSE = better prediction. Dashed line = overall model performance",
    x = "Baseline AIS Grade (Sample size)",
    y = "Root Mean Squared Error (RMSE)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40", margin = margin(b = 10)),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 0),
    axis.title = element_text(size = 11, face = "bold"),
    axis.title.x = element_text(margin = margin(t = 10)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ylim(0, max(subgroup_plot_data$rmse) * 1.25)

print(subgroup_plot)
```

## Clinical utility evaluation

We evaluate the model's clinical utility by defining a Minimal Clinically Important Difference (MCID) and plotting a histogram to find the percentage of predictions within this acceptable error.

```{r eval-util, fig.width=10, fig.height=7}

# Define Minimal Clinically Important Difference (MCID)
MCID <- 4

# Create binary outcomes based on MCID
clinical_data <- plot_data_cv %>%
  mutate(
    # Clinically acceptable prediction (within MCID)
    within_mcid = abs_residual <= MCID
  )

# Calculate proportion of predictions within MCID
within_mcid_pct <- mean(clinical_data$within_mcid) * 100

# Create error distribution plot
error_dist_plot <- ggplot(clinical_data, aes(x = abs_residual)) +
  geom_histogram(bins = 30, fill = "#2E86AB", alpha = 0.7, color = "white") +
  geom_vline(xintercept = MCID, color = "#E63946", 
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = MCID, y = Inf, 
           label = sprintf("MCID = %d", MCID),
           hjust = -0.1, vjust = 1.5, color = "#E63946", size = 4) +
  labs(
    title = "Distribution of prediction errors",
    subtitle = sprintf("%.1f%% of predictions are within MCID (±%d points)", within_mcid_pct, MCID),
    x = "Absolute prediction error (points)",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    axis.title = element_text(size = 11)
  )

error_dist_plot
```

We test the model's classification utility by setting a recovery threshold and generating a Precision-Recall curve from the model's predictions.

```{r eval-util-auprc, fig.width=10, fig.height=7}
# Define a clinically meaningful binary outcome
# This is a separate evaluation from the regression (MCID) utility
# Example: "Achieved a high recovery score" (e.g., > 50)
RECOVERY_THRESHOLD <- 60

# Create the binary 'truth' column
# We must use factors and set the "positive" class as the first level
class_data <- plot_data_cv %>%
  mutate(
    truth_class = factor(
      ifelse(motor_score_12m >= RECOVERY_THRESHOLD, "Recovered", "Not Recovered"),
      levels = c("Recovered", "Not Recovered")
    )
  )

# Calculate metrics
# We use `.pred` (our 0-100 score) as the ranking "probability"
# `yardstick` correctly handles this for ranking.
set.seed(123)
auroc_val <- roc_auc(class_data, truth = truth_class, .pred, event_level = "first")
auprc_val <- pr_auc(class_data, truth = truth_class, .pred, event_level = "first")
pr_curve_data <- pr_curve(class_data, truth = truth_class, .pred, event_level = "first")

# Calculate baseline prevalence for AUPRC plot
prevalence <- mean(class_data$truth_class == "Recovered")

# Plot PR Curve
auprc_plot <- ggplot(pr_curve_data, aes(x = recall, y = precision)) +
  geom_line(color = "#06A77D", linewidth = 1.2) +
  # Add baseline (no-skill) line
  geom_hline(yintercept = prevalence, linetype = "dashed", color = "gray40") +
  annotate("text", x = 0.5, y = prevalence + 0.05, 
           label = sprintf("Baseline (Prevalence) = %.2f", prevalence),
           color = "gray40") +
  labs(
    title = "Precision-Recall Curve (Classification Utility)",
    subtitle = sprintf("AUROC: %.3f | AUPRC: %.3f (Target: Score >= %d)", 
                       auroc_val$.estimate, auprc_val$.estimate, RECOVERY_THRESHOLD),
    x = "Recall (Sensitivity)",
    y = "Precision (PPV)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    axis.title = element_text(size = 11)
  )

auprc_plot
```

# **Permutation feature importance**

We conduct a permutation-based feature importance analysis, plotting the top 20 features ranked by the mean drop in performance (RMSE) when their values are shuffled.

```{r perm-global-importance}
# Create a custom predict function for the stacked model
predict_function_stack <- function(model, newdata) {
  pred <- predict(model, newdata)
  return(pred$.pred)
}

# Prepare data for explainer
# FIX: The 'data' argument must contain ALL original columns
# that the recipe was trained on, *except* the outcome.
# The recipe (inside the model) already knows which columns are IDs.
explain_data <- modeling_data %>%
  select(-motor_score_12m)

# The 'y' argument is just the outcome vector
explain_y <- modeling_data$motor_score_12m

# Create DALEX explainer
set.seed(123)
explainer_stack <- DALEX::explain(
  model = emsci_stack_fitted,
  data = explain_data,
  y = explain_y,
  predict_function = predict_function_stack,
  verbose = FALSE
)

print("DALEX explainer created successfully")

# Calculate permutation-based variable importance
# This will now work, as the 'explain_data' is complete.
set.seed(123)
tic("Permutation importance")
vi_permutation <- DALEX::model_parts(
  explainer_stack,
  loss_function = DALEX::loss_root_mean_square,
  B = 50,  # Number of permutations
  type = "difference"
)
toc()
```

```{r perm-global-importance-plot, fig.width=10, fig.height=8}
# Extract and prepare permutation importance data
perm_importance_data <- vi_permutation %>%
  as_tibble() %>%
  filter(variable != "_baseline_", variable != "_full_model_") %>%
  group_by(variable) %>%
  summarize(
    mean_dropout_loss = mean(dropout_loss),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_dropout_loss)) %>%
  slice_head(n = 20) %>%
  mutate(
    importance_category = case_when(
      mean_dropout_loss > quantile(mean_dropout_loss, 0.75) ~ "High",
      mean_dropout_loss > quantile(mean_dropout_loss, 0.25) ~ "Medium",
      TRUE ~ "Low"
    )
  )

# Create enhanced permutation importance plot
perm_plot <- ggplot(perm_importance_data, 
                    aes(x = reorder(variable, mean_dropout_loss), 
                        y = mean_dropout_loss,
                        fill = importance_category)) +
  geom_col(alpha = 0.85) +
  coord_flip() +
  # Color scale for importance levels
  scale_fill_manual(
    values = c("High" = "#E63946", "Medium" = "#F77F00", "Low" = "#FCBF49"),
    name = "Importance"
  ) +
  labs(
    title = "Global feature importance (permutation-based)",
    subtitle = "Top 20 features ranked by RMSE increase when feature values are randomly permuted 50 times. Higher = more important",
    x = "Feature",
    y = "Mean RMSE increase after permutation"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40", margin = margin(b = 10)),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 0),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text.y = element_text(size = 9),
    legend.position = "right",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

print(perm_plot)
```

# **SHAP feature importance**

We calculate SHapley Additive exPlanations (SHAP) values for the full dataset to explain the stacked model's predictions and create a global importance bar chart, ranking the top 20 features by their mean absolute SHAP value.

```{r shap-global-importance}
set.seed(123)

# For computational efficiency, use a sample of observations:
#sample_size <- min(500, nrow(explain_data))
#shap_sample_idx <- sample(1:nrow(explain_data), sample_size)
#shap_sample <- explain_data[shap_sample_idx, ]

shap_data <- explain_data
n_obs <- nrow(shap_data)

# Prediction wrapper
pfun <- function(object, newdata) {
  predict(object, newdata)$.pred
}

# Compute SHAP values using full dataset
set.seed(123)
tic("SHAP calculation (full sample)")
shap_values <- fastshap::explain(
  object = emsci_stack_fitted,
  X = shap_data,
  pred_wrapper = pfun,
  nsim = 50,
  adjust = TRUE
)
toc()

# Compute mean |SHAP|
shap_importance <- shap_values %>%
  as_tibble() %>%
  summarize(across(everything(), ~mean(abs(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "mean_abs_shap") %>%
  arrange(desc(mean_abs_shap)) %>%
  slice_head(n = 20)
```

```{r shap-global-importance-plot, fig.width=10, fig.height=8}
# --- Create shapviz object ---
# The shap_values and explain_data are from the previous chunk.

# Create the shapviz object. Using the explainer's y_intercept for the baseline (expected value).
# shapviz can consume the matrix output from fastshap::explain()
sv <- shapviz::shapviz(
  shap_values,
  X = shap_data, # Use the raw data for plotting feature values
  baseline = explainer_stack$y_intercept 
)

# 1. Global Importance Plot (Mean Absolute SHAP)
# Use sv_importance(kind="bar") which is equivalent to your original mean-absolute-SHAP bar chart.
shap_importance_plot <- sv_importance(sv, kind = "bar", show_top_n = 20) +
  labs(
    title = "Global Feature Importance (SHAP-based)",
    subtitle = sprintf("Top 20 features ranked by mean |SHAP| value (n=%d observations, nsim=50).", 
                      nrow(shap_data)),
    x = "Feature",
    y = "Mean Absolute SHAP Value (Average impact on prediction)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40", margin = margin(b = 10)),
    axis.title = element_text(size = 11, face = "bold")
  )

print(shap_importance_plot)
```

We generate SHAP dependency plots for the top numeric features to visualize their marginal effects and explore potential interactions.

```{r shap-dependency-plots, fig.width=12, fig.height=5, results = 'hide'}
# --- EXPERT IMPROVEMENT: SHAP Dependency Plots (No Patchwork) ---

# (patchwork library call removed)

# Ensure the sv object is available (it is created in the previous chunk)

# 1. Programmatically Identify Top Features (Robust selection)
top_features <- shap_importance %>% pull(feature)
top_numeric <- top_features[top_features %in% names(shap_data)[sapply(shap_data, is.numeric)]]
feat_marginal <- if (length(top_numeric) >= 1) top_numeric[1] else NULL
feat_interaction <- if (length(top_numeric) >= 2) top_numeric[2] else NULL


if (!is.null(feat_marginal)) {
  # Plot 1: Marginal Effect of the most important feature (non-linearity)
  dep_plot_marginal <- sv_dependence(
      sv, 
      v = feat_marginal, 
      alpha = 0.6
    ) +
    labs(
      title = paste("Marginal effect:", feat_marginal),
      subtitle = "SHAP value vs. feature value, colored by feature value."
    ) +
   theme_minimal() +
  theme(
    theme(legend.position = "bottom"),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    axis.title = element_text(size = 11)
  )

    
  # --- FIX: Print the first plot directly ---
  print(dep_plot_marginal)
    
  # Plot 2: Interaction Effect (using the second most important numeric feature for coloring)
  if (!is.null(feat_interaction) && feat_marginal != feat_interaction) {
    dep_plot_interaction <- sv_dependence(
        sv, 
        v = feat_marginal, 
        color_var = feat_interaction, # Color by a second numeric feature to visualize interaction
        alpha = 0.6
      ) +
      labs(
        title = paste("Interaction: Colored by", feat_interaction),
        subtitle = "Reveals how the effect is conditioned on the interaction feature."
      ) +
       theme_minimal() +
  theme(
    theme(legend.position = "bottom"),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    axis.title = element_text(size = 11)
  )

      
    # --- FIX: Print the second plot directly (will appear below the first) ---
    print(dep_plot_interaction)
    
  } else {
    message("Only one numeric feature found among the top features; displaying marginal plot only.")
  }
  
} else {
  message("No suitable numeric features found among the top features for dependency plots.")
}
```

# **Local Importance (Break-down)**

We identify the patients with the best and worst predictions from the cross-validation set and generate permutation-based "break-down" plots for these two individuals to visualize how each feature contributed to their specific final score.



```{r}
# --- SHAP-based Local Explanations for Best and Worst Predictions ---

# 1. ASSUME PRE-EXISTING OBJECTS from previous chunks:
#    - `plot_data_cv`: Contains OOF predictions and errors
#    - `shap_data`: The feature data used for SHAP
#    - `sv`: The 'shapviz' object created in 'shap-global-importance-plot'
#    - `explainer_stack`: The DALEX explainer (for baseline)

# 2. IDENTIFY BEST AND WORST PATIENTS (from OOF data)
#    (These objects are already in your Rmd 'perm-localexp' chunk)
best_patient_row <- plot_data_cv %>%
  arrange(abs_residual) %>%
  slice_head(n = 1)

worst_patient_row <- plot_data_cv %>%
  arrange(desc(abs_residual)) %>%
  slice_head(n = 1)

# 3. FIND ROW INDICES in the SHAP/In-Sample data
best_patient_id <- best_patient_row$patient_id_global
worst_patient_id <- worst_patient_row$patient_id_global

best_patient_index <- which(shap_data$patient_id_global == best_patient_id)
worst_patient_index <- which(shap_data$patient_id_global == worst_patient_id)

# 4. GET THE IN-SAMPLE PREDICTIONS (which the plots will explain)
#    The 'sv' object's prediction = baseline + sum of SHAP values
is_preds <- get_baseline(sv) + rowSums(get_shap_values(sv))
best_is_pred <- is_preds[best_patient_index]
worst_is_pred <- is_preds[worst_patient_index]

# 5. PLOT THE "BEST" PREDICTION
#    [cite_start]Using sv_waterfall from the shapviz package [cite: 1134, 1148]
set.seed(123)
plot_best_shap <- sv_waterfall(sv, row_id = best_patient_index) +
  labs(
    title = sprintf("SHAP Local Explanation: 'Best' OOF Prediction (Patient %s)", 
                    best_patient_id),
    subtitle = sprintf(
      "OOF: Obs=%.1f, Pred=%.1f, Error=%.2f  |  IN-SAMPLE: Pred=%.1f (This is plotted below)",
      best_patient_row$motor_score_12m,
      best_patient_row$.pred,
      best_patient_row$residual,
      best_is_pred
    )
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    axis.title = element_text(size = 11)
  )

print(plot_best_shap)

# 6. PLOT THE "WORST" PREDICTION
set.seed(123)
plot_worst_shap <- sv_waterfall(sv, row_id = worst_patient_index) +
  labs(
    title = sprintf("SHAP Local Explanation: 'Worst' OOF Prediction (Patient %s)", 
                    worst_patient_id),
    subtitle = sprintf(
      "OOF: Obs=%.1f, Pred=%.1f, Error=%.2f  |  IN-SAMPLE: Pred=%.1f (This is plotted below)",
      worst_patient_row$motor_score_12m,
      worst_patient_row$.pred,
      worst_patient_row$residual,
      worst_is_pred
    )
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    axis.title = element_text(size = 11)
  )

print(plot_worst_shap)
```
Note: OOF (Out-of-Sample) shows the model's unbiased prediction for this subject calculated during cross-validation, where the model had never seen this subject The OOF error is used only to identify this patient as "best" or "worst". IN-SAMPLE is the prediction from the final model, which was trained on all the data (including the chosen subjects). This is the prediction that the plot explains. The plot starts with the value for the average prediction for the dataset (E[f(x)]) and each bar is a feature's SHAP value, showing its push on the prediction until reaching the final prediction (f(x))

# **Limitations**

1. The model's performance is limited by its reliance on simulated data, which lacks the noise, heterogeneity, and missingness characteristic of real-world multi-center SCI data.

2. We implemented a standard two-stage CV approach. While it is computationally efficient and uses repeated grouped CV to obtain stable out-of-fold predictions, it carries an optimization bias. Nested CV would be preferred but computationally prohibitive.

3. The project uses embedded selection by passing all features to the stacked ensemble, employing the inherent feature selection capabilities of the base learners. Wrapper methods (e.g., Boruta) would be preferred but computationally prohibitive.

4. The analysis implements a pragmatic and computationally efficient two-stage evaluation in lieu of a full nested CV framework, introducing known trade-offs:

a. The hyperparameters for the base learners were selected using the same 5-fold CV that generated the out-of-fold predictions. This is standard for the stacks package but is slightly optimistic.

b. The optimal penalty for the meta-learner was selected in stage 1 using out-of-fold predictions from all 1500 patients. This single penalty was re-used in our stage 2. The final metrics are optimistic. This approach is a practical compromise that trades a minor, acceptable bias for a massive gain in computational efficiency.

5. Future iterations will include deeper investigation using SHAP dependency plots, partial dependence plots, and a quantitative analysis of the ensemble diversity.

# **Final summary**

```{r summary}
cat("These are not valid conclusions. This is a programming exercise. For completeness, here are the results:")
cat(sprintf("Overall performance (5-fold CV):"))
cat(sprintf("   - RMSE: %.2f (%.2f, %.2f)\n", 
            final_cv_metrics %>% filter(.metric == "rmse") %>% pull(mean),
            final_cv_metrics %>% filter(.metric == "rmse") %>% pull(mean) - 
              1.96 * final_cv_metrics %>% filter(.metric == "rmse") %>% pull(std_err),
            final_cv_metrics %>% filter(.metric == "rmse") %>% pull(mean) + 
              1.96 * final_cv_metrics %>% filter(.metric == "rmse") %>% pull(std_err)))
cat(sprintf("   - MAE: %.2f (%.2f, %.2f)\n", 
            final_cv_metrics %>% filter(.metric == "mae") %>% pull(mean),
            final_cv_metrics %>% filter(.metric == "mae") %>% pull(mean) - 
              1.96 * final_cv_metrics %>% filter(.metric == "mae") %>% pull(std_err),
            final_cv_metrics %>% filter(.metric == "mae") %>% pull(mean) + 
              1.96 * final_cv_metrics %>% filter(.metric == "mae") %>% pull(std_err)))
cat(sprintf("   - R-squared: %.3f (%.3f, %.3f)\n", 
            final_cv_metrics %>% filter(.metric == "rsq") %>% pull(mean),
            final_cv_metrics %>% filter(.metric == "rsq") %>% pull(mean) - 
              1.96 * final_cv_metrics %>% filter(.metric == "rsq") %>% pull(std_err),
            final_cv_metrics %>% filter(.metric == "rsq") %>% pull(mean) + 
              1.96 * final_cv_metrics %>% filter(.metric == "rsq") %>% pull(std_err)))

cat(sprintf("Calibration:"))
cat(sprintf("   - Calibration slope: %.3f (ideal = 1)\n", 
            calibration_metrics$calibration_slope))
cat(sprintf("   - Calibration intercept: %.3f (ideal = 0)\n", 
            calibration_metrics$calibration_intercept))

cat(sprintf("Clinical utility:"))
cat(sprintf("   - Predictions within MCID (±%d points): %.1f%%\n", 
            MCID, within_mcid_pct))

cat(sprintf("Bland-Altman analysis:"))
cat(sprintf("   - Mean bias: %.2f points\n", ba_stats$mean_diff))
cat(sprintf("   - 95%% Limits of Agreement: [%.2f, %.2f]\n", 
            ba_stats$loa_lower, ba_stats$loa_upper))

cat(sprintf("Meta-Learner composition:"))
cat(sprintf("   - Total members selected: %d (out of %d candidates)\n",
            nrow(meta_weights),
            ncol(emsci_stack) - 1)) # -1 for outcome
cat(sprintf("   - Dominant model type: %s (%.1f%% of total weight)\n",
            weights_by_model$model_type[1],
            100 * weights_by_model$total_weight[1] / sum(weights_by_model$total_weight)))

cat(sprintf("Subgroup performance:"))
for (ais in c("A", "B", "C", "D")) {
  if (ais %in% subgroup_metrics$baseline_ais) {
    metrics <- subgroup_metrics %>% filter(baseline_ais == ais)
    cat(sprintf("   - AIS %s (n=%d): RMSE=%.2f, R-squared=%.3f\n",
                ais, metrics$n, metrics$rmse, metrics$rsq))
  }
}
```

## Session Information

```{r session-info, echo=FALSE}
cat("R version:", R.version$version.string, "\n")
cat("Platform:", R.version$platform, "\n\n")
cat("Key ML packages:\n")
pkgs <- c("tidymodels", "tune", "stacks", 
          "ranger", "xgboost", "earth", "DALEX", "fastshap")
for(pkg in pkgs) {
  if(requireNamespace(pkg, quietly = TRUE)) {
    cat(pkg, as.character(packageVersion(pkg)), "\n")
  }
}
```

